---
title: "Equity Report: Child Poverty"
author: "[UVA CommPAS Lab](http://commpas-lab.mystrikingly.com/)"
date: "Hannah Lewis and Michele Claibourn"
output: 
  html_document:
    theme: spacelab
    highlight: tango
    toc: true
    toc_depth: 2
---

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri(file.path(getwd(), "image", "big-compass.png")), 
               alt = 'commpas', 
               style = 'position:absolute; top:0; right:0; padding:20px; height:225px')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r loading}
library(tidyverse)
library(knitr)
library(kableExtra)
library(formattable)
library(leaflet)
library(RColorBrewer)
library(pander)

state <- "VA"
county <- "003"
year <- 2017

# Read in county codes table
county_codes <- read_csv("county_codes.csv")
row <- which(county_codes$code == county)

filename <- paste("cPov_",county,"_tract_",year,".RDS", sep = "")
filename_geo <- paste("cPov_",county,"_tract_",year,"_geo",".RDS", sep = "")

# Load equity table
equityTable_cPov <- readRDS(filename) 
equityTable_cPov_geo <- readRDS(filename_geo) 
```
***
### About the Indicator Reports
The following report presents an equity indicator, inspired by efforts like that of [St. Louis Equity Indicators Project](https://www.stlouis-mo.gov/government/departments/mayor/initiatives/resilience/equity/index.cfm) emphasizing racial or ethnic disparities in measures of community well-being. The goal is to make indicators from sources like the U.S. Census American Communit Survey readily accessible and reproducible. Indicators derived from sampled data, like the American Community Survey are measured with uncertainty and we aim to make that uncertainy apparent.

The prototype report focuses on child poverty and contains three sections: 

* Regional summaries, providing an explanation of the indicator, a summary table showing outcomes for selected population groups in the region along with disparity ratios, and details about the data source; 
* Maps by census tract, mapping the indicator for each selected population, to help reveal variations within the locality;
* Population and rate tables, showing the population counts and indicator for each selected population group by tract along witn the margins of error to make plain the uncertainty around some estimates at this smaller geographic level. The indicator values from this table are the values represented in the maps. 

***

# **Child Poverty in `r county_codes[row, 3]`**

### What does this indicator measure
Child poverty measures the percentage of children under age 18 who live in households with incomes at or below the federal poverty line in `r county_codes[row, 3]`. The federal poverty line was \$24,600 for a family of four in 2017. In `r year`, there were `r last(equityTable_cPov$cPov)` (+/- `r round(last(equityTable_cPov$cPov_moe),0)`) children living in poverty in `r county_codes[row, 3]`, `r round(last(equityTable_cPov$Per),1)`% of all children. 

***
## Summary Table 
Children under age 18 who live in poverty in `r county_codes[row, 3]`

|             | Total Children | White Children | Black Children | Hispanic Children | Black-White Disparity | Hispanic-White Disparity | 
|------------|:-----:|:-----:|:-----:|:-----:|:-----:|:-----:|
| **Children in poverty** | `r last(equityTable_cPov$cPov)` | `r last(equityTable_cPov$cPov_A)` | `r last(equityTable_cPov$cPov_B)` | `r last(equityTable_cPov$cPov_C)` |  |  | 
| **Child population**    | `r last(equityTable_cPov$pop)`  | `r last(equityTable_cPov$pop_A)` | `r last(equityTable_cPov$pop_B)` | `r last(equityTable_cPov$pop_C)` |  |  |  
| **Child poverty rate**  | `r round(last(equityTable_cPov$Per),1)`% | `r round(last(equityTable_cPov$Per_A),1)`% | `r round(last(equityTable_cPov$Per_B),1)`% | `r round(last(equityTable_cPov$Per_C),1)`% | `r round(last(equityTable_cPov$Ratio_B),1)` | `r round(last(equityTable_cPov$Ratio_C),1)` | 

**Data Source:** U.S. Census Bureau; American Community Survey, 2013-2017, derived from Tables B01001, B01001H, B01001B, B01001I, B17001, B17001H, B17001B, and B17001I.

**Data Explanation:** Local child population and children in poverty estimates are from the American Community Survey 5-Year Estimates, 2013-2017. Sent to approximately 3.5 million addresses per year, the 5-year survey estimates provide up-to-date estimates for localities that may be changing between censuses. As these are estimates derived from surveys, and thus subject to variability due to sampling error, margins of error are included in the tables below. 

***

### Disparity Ratio 
The disparity ratio is the percentage of black or Hispanic children living in poverty divided by the percentage of white children living in poverty. A disparity ratio of 1 reflects racial equality; values greater than one suggest higher rates of poverty among black or Hispanic children compared to white children while values less than 1 suggest lower rates of poverty among black or Hispanic children. 

* In `r county_codes[row, 3]`, black children experience poverty at notably greater rates than do white children. The disparity ratio shows that black children are `r round(last(equityTable_cPov$Ratio_B),1)` times more likely to live in households with incomes at or below the federal poverty line. 
* In `r county_codes[row, 3]`, Hispanic children experience poverty at notably greater rates than do white children. The disparity ratio shows that Hispanic children are `r round(last(equityTable_cPov$Ratio_C),1)` times more likely to live in households with incomes at or below the federal poverty line.


***

## Child Poverty by Census Tract

Within each map, clicking on a tract will show the estimated child population, the estimated children in poverty, and the estimated poverty rate. For child poverty, lighter colors show lower poverty rates and darker colors show higher poverty rates. 


**Child Poverty among Black Children**

```{r map1, out.width = '100%'}
# child poverty-black
# define popup info
popup <- paste0("Child Population: ", equityTable_cPov_geo$pop_B,
                "<br>", "Children in Poverty: ", equityTable_cPov_geo$cPov_B,
                "<br>", "Child Poverty Rate: ", round(equityTable_cPov_geo$Per_B, 2),
                "<br>", "Tract: ", equityTable_cPov_geo$NAME)

min <- min(c(equityTable_cPov_geo$Per_B, equityTable_cPov_geo$Per_A), na.rm = TRUE)
max <- max(c(equityTable_cPov_geo$Per_B, equityTable_cPov_geo$Per_A), na.rm = TRUE)+1

# define palette
# Define the number of colors you want
nb.cols <- 20
mycolors <- colorRampPalette(brewer.pal(8, "YlGnBu"))(nb.cols)

pal <- colorNumeric(palette = mycolors,
                    domain = min:max)

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  # addTiles() %>% # to show streets more prominently
  addPolygons(data = equityTable_cPov_geo,
              fillColor = ~pal(equityTable_cPov_geo$Per_B),
              fillOpacity = 0.5,
              color = "white",
              weight = 2,
              smoothFactor = 0.2,
              popup = popup,
              highlight = highlightOptions(
                weight = 5,
                fillOpacity = 0.7,
                bringToFront = TRUE)) %>%
  addLegend(pal = pal,
            values = equityTable_cPov_geo$Per_B,
            position = "bottomright",
            opacity = 0.5,
            title = paste("Child Poverty Rate<br>among Black Children<br>in ",year, sep = ""))
```

NOTE: Currently any descriptive text is added manually after examining the output.

**Child Poverty among Hispanic Children**

```{r map2, out.width = '100%'}
# child poverty-Hispanic
# define popup info
popup <- paste0("Child Population: ", equityTable_cPov_geo$pop_C,
                "<br>", "Children in Poverty: ", equityTable_cPov_geo$cPov_C,
                "<br>", "Child Poverty Rate: ", round(equityTable_cPov_geo$Per_C, 2),
                "<br>", "Tract: ", equityTable_cPov_geo$NAME)

min <- min(c(equityTable_cPov_geo$Per_C, equityTable_cPov_geo$Per_A), na.rm = TRUE)
max <- max(c(equityTable_cPov_geo$Per_C, equityTable_cPov_geo$Per_A), na.rm = TRUE)+1

# define palette
# Define the number of colors you want
nb.cols <- 20
mycolors <- colorRampPalette(brewer.pal(8, "YlGnBu"))(nb.cols)

pal <- colorNumeric(palette = mycolors,
                    domain = min:max)

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  # addTiles() %>% # to show streets more prominently
  addPolygons(data = equityTable_cPov_geo,
              fillColor = ~pal(equityTable_cPov_geo$Per_C),
              fillOpacity = 0.5,
              color = "white",
              weight = 2,
              smoothFactor = 0.2,
              popup = popup,
              highlight = highlightOptions(
                weight = 5,
                fillOpacity = 0.7,
                bringToFront = TRUE)) %>%
  addLegend(pal = pal,
            values = equityTable_cPov_geo$Per_C,
            position = "bottomright",
            opacity = 0.5,
            title = paste("Child Poverty Rate<br>among Hispanic Children<br>in ",year, sep = ""))
```

NOTE: Currently any descriptive text is added manually after examining the output.


**Child Poverty among White Children**

```{r map3, out.width = '100%'}
# child poverty-white
# define popup info
popup <- paste0("Child Population: ", equityTable_cPov_geo$pop_A,
                "<br>", "Children in Poverty: ", equityTable_cPov_geo$cPov_A,
                "<br>", "Child Poverty Rate: ", round(equityTable_cPov_geo$Per_A, 2),
                "<br>", "Tract: ", equityTable_cPov_geo$NAME)
# define palette
pal <- colorNumeric(palette = mycolors,
                    domain = min:max)

leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  # addTiles() %>% # to show streets more prominently
  addPolygons(data = equityTable_cPov_geo,
              fillColor = ~pal(equityTable_cPov_geo$Per_A),
              fillOpacity = 0.5,
              color = "white",
              weight = 2,
              smoothFactor = 0.2,
              popup = popup,
              highlight = highlightOptions(
                weight = 5,
                fillOpacity = 0.7,
                bringToFront = TRUE)) %>%
  addLegend(pal = pal,
            values = equityTable_cPov_geo$Per_A,
            position = "bottomright",
            opacity = 0.5,
            title = paste("Child Poverty Rate<br>among White Children<br>in ",year, sep = ""))
```

NOTE: Currently any descriptive text is added manually after examining the output.


***
## Data Tables
The following tables show the population estimates and poverty rates for each census tract. The tables also provide the margains of error associated with each estimate. ACS data is based on a sample and so estimates contain a degree of uncertainty. Larger margins mean the estimate is less accurate. Typically the larger the sample size, the smaller the margin of error. This means in cases where the population of interest is very small (very few children live in a census tract), estimates can be very unreliable. 

### Missing values
In the poverty rate table, NaN's exist when there is division by zero, that is, when the population of interest in that census tract is estimated to be zero. Poverty rate estimates of zero occur when the number of children estimated to be in poverty is zero.

***

**Population Estimates Table**

```{r poptable}
tbl <- equityTable_cPov %>%
  select(NAME:pop_moe_C) %>%
  mutate(Tract = str_remove(NAME, paste(", ", county_codes[row,3], ", Virginia", sep="")),
         Children = paste(pop, "(", "+/-", round(pop_moe, 0), ")"),
         White = paste(pop_A, "(", "+/-", round(pop_moe_A, 0), ")"),
         Black = paste(pop_B, "(", "+/-", round(pop_moe_B, 0), ")"),
         Hispanic = paste(pop_C, "(", "+/-", round(pop_moe_C, 0), ")")) %>%
  select(Tract, Children, White, Black, Hispanic) %>%
  mutate(Tract = str_remove(Tract, ", Virginia"))

kable(tbl, align = c('l','c','c','c','c'), digits = 0) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
     footnote(general = "Margins of error are shown in parentheses and indicate that the population could be more or less than the estimate by the value of the margin of error.")
```

**Poverty Rate Estimates Table**

```{r povtable}
tbl <- equityTable_cPov %>% select(NAME,Per:Per_moe_C) %>%
  mutate(Tract = str_remove(NAME, paste(", ", county_codes[row,3], ", Virginia", sep="")),
         `Pov Rate` = paste(round(Per,1), "(", "+/-", round(Per_moe,1), ")"),
         `Pov Rate-White` = paste(round(Per_A,1), "(", "+/-", round(Per_moe_A,1), ")"),
         `Pov Rate-Black` = paste(round(Per_B,1), "(", "+/-", round(Per_moe_B,1), ")"),
         `Pov Rate-Hispanic` = paste(round(Per_C,1), "(", "+/-", round(Per_moe_C,1), ")")) %>%
  select(Tract, `Pov Rate`, `Pov Rate-White`, `Pov Rate-Black`, `Pov Rate-Hispanic`) %>%
  mutate(Tract = str_remove(Tract, ", Virginia"))

kable(tbl, align = c('l', 'c', 'c','c','c'), digits = 1) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
     footnote(general = "Margins of error are shown in parentheses and indicate that the poverty rates could be more or less than the estimate by the value of the margin of error.")
```
